### 2019-01-28 treestats_paper.tex

\newcommand{\plr}[1]{{\color{blue} \it #1}}

@@

The key to the remarkable scalability and efficiency of tree sequence
algorithms is the way in which the shared structure in adjacent trees along
the genome is encoded. As one looks across the genome the genealogies change
because of the results of recombination in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be used
to both store genome sequence and the associated genealogical relationships extremely compactly,
as well as to compute statistics highly efficiently~\citet{kelleher2016efficient}.
This paper generalizes the latter strategy to a much broader class of statistics.

@@

% summary
In this paper, we present a general framework for defining and efficiently computing
``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site.
Patterns of genetic polymorphism in a population
are determined by the population's history of Mendelian inheritance,
as summarized by the underlying genealogical trees,
and so it is natural (and helpful) to express these statistics
in terms of the structure of the trees themselves.
This perspective turns out to both suggest efficient algorithms
for computing these statistics on tree sequences,
and allows us to treat single-site genetic statistics in a more general framework
of summaries of tree shape and inheritance.
The methods are implemented in the python package \tskit.

@@

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome.
Suppose first that we wish to compute an additive function of the allele frequency spectrum.
In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
and that the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
This definition applies to multiallelic sites by assuming that for each allele
we can only see how many samples carry that allele,
not how the remaining samples are partitioned between one or more other alleles.
(Note also that this does not differentiate ancestral and derived alleles,
and so is a function of the \emph{folded} allele frequency spectrum.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and removing those that subsequently mutate).

There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
and so a general statistic of biallelic genotypes could be represented by a vector of length $2^n$.
A more convenient way to parameterize many functions of genotype patterns
uses three ingredients:
\begin{itemize}

    \item[(a)] A list of $K$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$,
        where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.

@@

    \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that
        $F(\nw)$ is the ``summary'' statistic we compute

\end{itemize}
Roughly speaking, we use these to compute a statistic
by first propagating the initial weights up the trees,
and then evaluate the summary function on the resulting vectors of weights
for each allele.

>>>>

\newcommand{\plr}[1]{{\color{blue}\textbf{plr:} \it #1}}
\newcommand{\jk}[1]{{\color{red}\textbf{jk:} \it #1}}

@@

The key to the remarkable scalability and efficiency of tree sequence
algorithms is the way in which the shared structure in adjacent trees along
the genome is encoded. As one looks across the genome the genealogies change
because of the results of recombination in ancestors.
However, nearby trees tend to share a lot of common structure,
and single genealogical relationships (e.g., individual $x$ inherits from individual $y$)
are often shared across relatively long distances,
which manifest as shared \emph{edges} across many adjacent trees in the tree sequence.
This redundancy can be used
to both store genome sequence and the associated genealogical relationships extremely compactly,
as well as to compute statistics highly efficiently~\citep{kelleher2016efficient}.
This paper generalizes the latter strategy to a much broader class of statistics.

@@

% summary
In this paper, we present a general framework for defining and efficiently computing
``single-site'' genetic statistics,
i.e., statistics of aligned genome sequence that can be expressed as averages over values computed
separately for each site.
Patterns of genetic polymorphism in a population
are determined by the population's history of Mendelian inheritance,
as summarized by the underlying genealogical trees,
and so it is natural (and helpful) to express these statistics
in terms of the structure of the trees themselves.
This perspective turns out to both suggest efficient algorithms
for computing these statistics on tree sequences,
and allows us to treat single-site genetic statistics in a more general framework
of summaries of tree shape and inheritance.
The methods are implemented in the \tskit\ library.

@@

Suppose we have genotyped a collection of $N$ (homologous) chromosomes
at a sequence of $L$ loci along the genome.
\jk{Maybe use $n$ here? We use it below}
Suppose first that we wish to compute an additive function of the allele frequency spectrum.
In other words, suppose that the frequency of allele $a$ at a given site is $p_i(a)$,
and that the allele frequency spectrum
-- i.e., the proportion of alleles whose frequency is $p$ in our sample -- is $f(p)$,
and for some function $h()$ we want to compute $\sum_p f(p) h(p)$.
Since this is a simple average over sites,
we can compute this, summing over the $L$ sites, as
$\sum_p f(p) h(p) = (1/L) \sum_{i=1}^L \sum_a h(p_i(a))$.
This definition applies to multiallelic sites by assuming that for each allele
we can only see how many samples carry that allele,
not how the remaining samples are partitioned between one or more other alleles.
\jk{Not clear to me here if we're talking about frequency $0 \leq p \leq 1$ or
`prevalance`, $0 \leq p \leq N$.}
(Note also that this does not differentiate ancestral and derived alleles,
and so is a function of the \emph{folded} allele frequency spectrum.)
If we know where each mutation has occurred on the genealogy,
we can easily find the frequency of the corresponding allele by counting how many samples
occur below that mutation in the tree (and removing those that subsequently mutate).

\jk{Big jump here. I don't see the connection between the previous paragraph
and the bitstring representation of haplotype statistics}
There are $2^n$ possible genotype patterns for $n$ haploid samples genotyped at a biallelic locus,
and so a general statistic of biallelic genotypes could be represented by a vector of length $2^n$.
A more convenient way to parameterize many functions of genotype patterns
uses three ingredients:
\begin{itemize}

    \item[(a)] A list of $k$ vectors of \emph{initial node weights}, $(\iw_1, \ldots, \iw_k)$,
        where $\iw_i(j)$ is the ``initial'' value of the $i^\text{th}$ weight assigned to node $j$.
        \jk{Maybe referring to nodes with variables $u, v, w$ and keeping $i,
        j, k$ for general indexes would help here separate the concepts here?}

@@

    \item[(c)] A \emph{summary function} $F : \R^k \to \R$ so that
        $F(\nw)$ is the ``summary'' statistic we compute \jk{I'd find it more
        intuitive to use $f$ as the notation for a summary function. We can use
        something else to define the site frequency spectrum above, right? Big
        $F$ says ``set" to me.}

\jk{This formulation of a $k$-vector associated with each node is very
reminiscent of classical phlogenetics algorithms. Eg the Sankoff algorithm
for counting the number of mutations needed to explain a set of genotypes
under parsimony and with arbitrary weighting matrices. See
\citep[pg.13]{felsenstein2004inferring}. We should emphasise this connection.
Perhaps a diagram would help here also, showing a tree and a $k$-vector on each
node?}

Roughly speaking, we use these to compute a statistic
by first propagating the initial weights up the trees,
and then evaluate the summary function on the resulting vectors of weights
for each allele.
